---
title: "Lunaphore Cell Quantification - Tumor Samples"
output: html_document
params:
  sample_name: "ACCX"
  input_file: "ACCX_cell_intensities.csv"
  thresholds_file: "thresholds.csv"
  results_outdir: "quantification"
---

<style type="text/css">
.main-container {
max-width: 1600px;
margin-left: 20px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
# knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)  # suppress for clean html

library(dplyr)
library(ggplot2)
library(Seurat)
library(knitr)
library(UpSetR)
```

## `r params$sample_name`

```{r}
# Prep input intensity data
df <- data.table::fread(params$input_file) %>% as.data.frame()

df <- df[!is.na(rowMeans(df %>% select(ends_with("Mean")))),]  # remove cells with failed segmentation

# Fix column name inconsistency
colnames(df) <- gsub(" \\(C\\d*\\)", "", colnames(df))
colnames(df) <- gsub("CD-45", "CD45", colnames(df)) 
colnames(df) <- gsub("cMYB", "c-Myb", colnames(df))
colnames(df) <- gsub("T_PDL1", "PD-L1", colnames(df))
colnames(df) <- gsub("PDL2", "PD-L2", colnames(df))
colnames(df) <- gsub("T_Ki67", "Ki67", colnames(df))
colnames(df) <- gsub("HLA-ABC", "HLA", colnames(df))
colnames(df) <- gsub("PD1", "PD-1", colnames(df))

df <- df[, !grepl("TRITC - TRITC", colnames(df))]

rownames(df) <- df$`Object ID`

# Remove annotated artifacts if present
if ("is_artifact" %in% colnames(df)) {
  df <- df %>% filter(!is_artifact)
}

df$Parent <- ifelse(df$Parent == "Root object (Image)", "non_tumor", df$Parent)  # rename non-tumor section
table(df$Parent) %>% kable(col.names=c("Annotation", "# Cells"))

# Prep threshold data for specified sample
thresh_df <- read.csv(params$thresholds_file)
thresh_df <- thresh_df %>% select(marker, params$sample_name)
colnames(thresh_df) <- c("marker", "manual")
```

#### Shape Features

```{r}
area_dat <- df %>% select(matches("^Nucleus|^Cell")) %>% filter(!is.na(df$`DAPI: Cytoplasm: Mean`))
t(summary(area_dat)) %>% knitr::kable()
```

```{r, fig.width=12, fig.height=8}
dat <- area_dat %>% mutate(Object.ID=rownames(area_dat))
dat <- dat %>% tidyr::pivot_longer(!Object.ID, names_to="Feature", values_to="Value")
ggplot(dat, aes(x=Value)) +
  geom_histogram(bins=1000, fill="steelblue", alpha=0.8) +
  facet_wrap(vars(Feature), ncol=5, scales="free") +
  theme_bw()
```

#### Intensity Features

```{r}
intensity_dat <- df %>% select(ends_with("Mean")) %>% filter(!is.na(df$`DAPI: Cytoplasm: Mean`))
t(summary(intensity_dat)) %>% knitr::kable()
```

#### Nuclear/Cytoplasmic Ratios for Each Marker

```{r, fig.width=12, fig.height=8}
ratio_dat <- data.frame(Object.ID=rownames(intensity_dat))
markers <- unique(gsub(":.*", "", colnames(intensity_dat)))
for (marker in markers) {
  ratio_dat[[marker]] <- intensity_dat[[paste0(marker, ": Nucleus: Mean")]]/intensity_dat[[paste0(marker, ": Cytoplasm: Mean")]]
}

dat <- ratio_dat %>% tidyr::pivot_longer(!Object.ID, names_to="MarkerID", values_to="Ratio") %>% mutate(Marker=gsub("\\:.*", "", MarkerID))
ggplot(dat, aes(x=log2(Ratio))) +
  geom_histogram(bins=100, fill="steelblue", alpha=0.8) +
  facet_wrap(vars(Marker), ncol=5, scales="free") +
  theme_bw()
```

```{r}
markers <- unique(gsub(":.*", "", colnames(intensity_dat)))
nuc_markers <- grep("Myb|p63|NFIB|SOX2|FoxP3|Ki67", markers, value=TRUE)
auto_markers <- c("Cy5", "TRITC")
cyto_markers <- setdiff(markers, c(nuc_markers, auto_markers, "DAPI"))

mean_dat <- df %>% select(`DAPI: Nucleus: Mean`, paste0(c(auto_markers, cyto_markers), ": Cell: Mean"), paste0(nuc_markers, ": Nucleus: Mean"))
colnames(mean_dat) <- gsub("(:)? .*", "", colnames(mean_dat))
markers <- colnames(mean_dat)
```

### Intensity Histograms with Thresholds Marked

```{r, fig.width=12, fig.height=8}
mean_cols <- c("DAPI: Nucleus: Mean", paste0(c(auto_markers, cyto_markers), ": Cell: Mean"), paste0(nuc_markers, ": Nucleus: Mean"))
dat <- intensity_dat %>% select(all_of(mean_cols)) %>% mutate(Object.ID=rownames(intensity_dat))
dat <- dat %>% tidyr::pivot_longer(all_of(mean_cols), names_to="MarkerID", values_to="Intensity")
dat <- dat %>% mutate(marker=gsub("(:)? .*", "", MarkerID), marker_channel=gsub("\\:.*", "", MarkerID))
dat <- merge(dat, thresh_df, by="marker", all.x=TRUE)
dat$marker_channel <- factor(dat$marker_channel, levels=c("DAPI", auto_markers, cyto_markers, nuc_markers))

ggplot(dat, aes(x=log2(Intensity))) +
  geom_histogram(bins=100, fill="steelblue", alpha=0.8) +
  geom_vline(aes(xintercept=log2(manual)), color="firebrick") +
  facet_wrap(vars(marker_channel), ncol=5, scales="free") +
  theme_bw()
```

### Plot All Cells by Image Centroid Coordinates

Blue: DAPI, Green: CK

```{r, fig.width=12, fig.height=12}
min_max_scale <- function(col) { (col - min(col))/max(col) }

dat <- df %>% select(`Object ID`, `Centroid X µm`, `Centroid Y µm`, `CK - TRITC: Cell: Mean`, `DAPI: Nucleus: Mean`)
dat$DAPI_scaled <- min_max_scale(dat$`DAPI: Nucleus: Mean` + 1)
dat$CK_scaled <- min_max_scale(dat$`CK - TRITC: Cell: Mean` + 1)

ggplot(dat, aes(x=`Centroid X µm`, y=-`Centroid Y µm`)) +
  geom_point(size=0.1, color="#0345fc", aes(alpha=DAPI_scaled)) +
  geom_point(size=0.1, color="#03fc77", aes(alpha=CK_scaled)) +
  guides(alpha="none") +
  theme_minimal()

ggplot(df %>% mutate(is_tumor=ifelse(is_tumor, "Tumor", "Non-Tumor")), aes(x=`Centroid X µm`, y=-`Centroid Y µm`, color=is_tumor)) +
  geom_point(size=0.1, alpha=0.5) +
  theme_minimal() +
  scale_color_manual(values=c("Tumor"="#C23637", "Non-Tumor"="#0068A9")) +
  # guides(color="none") +
  guides(color=guide_legend(override.aes=list(size=3))) +
  labs(color="Annotation")
```

### Filter Data

```{r}
paste("Original Cells:", nrow(df))
dapi_thresh <- thresh_df$manual[thresh_df$marker == "DAPI"]
low_dapi <- df$`DAPI: Nucleus: Mean` < dapi_thresh  # low DAPI intensity
print(paste("Low DAPI:", sum(low_dapi)))
cyto_dapi <- df$`DAPI: Nucleus: Mean`/df$`DAPI: Cytoplasm: Mean` < 1  # DAPI not nuclear
print(paste("Non-Nuclear DAPI:", sum(cyto_dapi)))
mean_num_zeros <- apply(df %>% select(all_of(mean_cols)), 1, function(row) sum(row == 0))
zero_mean <- mean_num_zeros > 0  # extreme outlier cells (contain zero intensity values)
print(paste("Contains Zero:", sum(zero_mean)))
extreme_outlier <- rowSums(scale(log2(df %>% select(all_of(mean_cols)))) > 10, na.rm=TRUE) > 0  # extreme outlier cells (more than 10 sd above mean) 
print(paste("Extreme Value Outliers:", sum(extreme_outlier)))
discard <- low_dapi | cyto_dapi | zero_mean | extreme_outlier
paste("Discarded Cells:", sum(discard))
```

#### Plot Discarded Cells by Image Centroid Coordinates

```{r, fig.width=12, fig.height=12}
dat <- df[discard,] %>% select(`Object ID`, `Centroid X µm`, `Centroid Y µm`, `CK - TRITC: Cell: Mean`, `DAPI: Nucleus: Mean`)
dat$DAPI_scaled <- min_max_scale(dat$`DAPI: Nucleus: Mean` + 1)
dat$CK_scaled <- min_max_scale(dat$`CK - TRITC: Cell: Mean` + 1)

ggplot(dat, aes(x=`Centroid X µm`, y=-`Centroid Y µm`, alpha=CK_scaled)) +
  geom_point(size=0.1, color="#0345fc", aes(alpha=DAPI_scaled)) +
  geom_point(size=0.1, color="#03fc77", aes(alpha=CK_scaled)) +
  guides(alpha="none") +
  theme_minimal() +
  labs(alpha="CK Scaled")
```

#### Plot Retained Cells by Image Centroid Coordinates

```{r, fig.width=12, fig.height=12}
df <- df[!discard,]
mean_dat <- mean_dat[!discard,]
paste("Retained Cells:", nrow(df))

dat <- df %>% select(`Object ID`, `Centroid X µm`, `Centroid Y µm`, `CK - TRITC: Cell: Mean`, `DAPI: Nucleus: Mean`)
dat$DAPI_scaled <- min_max_scale(dat$`DAPI: Nucleus: Mean` + 1)
dat$CK_scaled <- min_max_scale(dat$`CK - TRITC: Cell: Mean` + 1)

ggplot(dat, aes(x=`Centroid X µm`, y=-`Centroid Y µm`, alpha=CK_scaled)) +
  geom_point(size=0.1, color="#0345fc", aes(alpha=DAPI_scaled)) +
  geom_point(size=0.1, color="#03fc77", aes(alpha=CK_scaled)) +
  guides(alpha="none") +
  theme_minimal() +
  labs(alpha="CK Scaled")
```

## Classify Cell Types by Thresholds

```{r}
mean_dat <- mean_dat %>% select(!c(DAPI, Cy5, TRITC))
markers <- colnames(mean_dat)

classification_markers <- c("CD3", "CD4", "CD8", "CD68", "CD20", "CK", "Ki67", "PD-1", "FoxP3")
thresholded_markers <- thresh_df$marker[!is.na(thresh_df$manual)]
stopifnot(length(setdiff(markers, thresholded_markers)) == 0)  # all markers should have thresholds set

# Apply binary threshold
thresh_dat <- lapply(markers, function(marker) {
  col <- mean_dat[marker]
  thresh <- thresh_df$manual[thresh_df$marker == marker]
  col[col < thresh] <- 0
  col[col >= thresh] <- 1
  col
}) %>% as.data.frame(check.names=FALSE)
```

#### Upset Plot of Top 30 Positive Marker Combinations

```{r, fig.width=12, fig.height=6}
upset_dat <- lapply(classification_markers, function(marker) rownames(thresh_dat)[thresh_dat[[marker]] == 1])
names(upset_dat) <- classification_markers

upset(fromList(upset_dat), number.angles=30, point.size=2, line.size=1, text.scale=1.2, nsets=length(upset_dat), 
      nintersects=30, order.by="freq", main.bar.color="#004586", sets.bar.color="#018F99")
```

```{r}
cell_meta <- data.frame(cell_id=rownames(thresh_dat), annotation=df$Parent, is_tumor=df$is_tumor, is_gland=df$is_gland)

cell_meta$pdl1_pos <- thresh_dat$`PD-L1` == 1

if ("PD-L2" %in% colnames(thresh_dat)) {
  cell_meta$pdl2_pos <- thresh_dat$`PD-L2` == 1
}

thresh_dat <- thresh_dat %>% select(all_of(classification_markers))

cell_meta$pos_markers <- apply(thresh_dat, 1, function(row) paste(names(row)[row == 1][order(names(row)[row == 1])], collapse=", "))

cell_meta <- cell_meta %>% 
  mutate(cell_subtype=case_when(pos_markers %in% c("CK", "CK, PD-1") ~ "Epithelial",
                                pos_markers %in% c("CK, Ki67", "CK, Ki67, PD-1") ~ "Epithelial: Proliferating",
                                pos_markers == "CD3, CD4" ~ "T-Helper",
                                pos_markers == "CD3, CD4, CK" ~ "T-Helper: on Epithelial",
                                pos_markers == "CD3, CD4, FoxP3" ~ "T-Helper: Regulatory",
                                pos_markers == "CD3, CD4, CK, FoxP3" ~ "T-Helper: Regulatory/on Epithelial",
                                pos_markers == "CD3, CD8" ~ "Cytotoxic T",
                                pos_markers == "CD3, CD8, CK" ~ "Cytotoxic T: on Epithelial",
                                pos_markers == "CD3, CD8, Ki67" ~ "Cytotoxic T: Proliferating",
                                pos_markers == "CD3, CD8, CK, Ki67" ~ "Cytotoxic T: Proliferating/on Epithelial",
                                pos_markers == "CD3, CD8, PD-1" ~ "Cytotoxic T: Exhausted",
                                pos_markers == "CD3, CD8, CK, PD-1" ~ "Cytotoxic T: Exhausted/on Epithelial",
                                pos_markers %in% c("CD68", "CD4, CD68") ~ "Macrophage",
                                pos_markers %in% c("CD68, CK", "CD4, CD68, CK") ~ "Macrophage: on Epithelial",
                                pos_markers %in% c("CD68, Ki67", "CD4, CD68, Ki67") ~ "Macrophage: Proliferating",
                                pos_markers %in% c("CD68, CK, Ki67", "CD4, CD68, CK, Ki67") ~ "Macrophage: Proliferating/on Epithelial",
                                pos_markers %in% c("CD68, PD-1", "CD4, CD68, PD-1") ~ "Macrophage: Suppressed",
                                pos_markers %in% c("CD68, CK, PD-1", "CD4, CD68, CK, PD-1") ~ "Macrophage: Suppressed/on Epithelial",
                                pos_markers == "CD20" ~ "B Cell",
                                pos_markers == "CD20, CK" ~ "B Cell: on Epithelial",
                                pos_markers == "CD20, Ki67" ~ "B Cell: Proliferating",
                                pos_markers == "CD20, CK, Ki67" ~ "B Cell: Proliferating/on Epithelial",
                                TRUE ~ "Unknown"))

cell_meta <- cell_meta %>% mutate(cell_type=gsub(":.*", "", cell_subtype))
```

#### Upset Plot of All Valid Positive Marker Combinations

```{r, fig.width=12, fig.height=6}
classified_cells <- cell_meta$cell_id[cell_meta$cell_type != "Unknown"]
classified_thresh_dat <- thresh_dat[rownames(thresh_dat) %in% classified_cells,]
upset_dat <- lapply(classification_markers, function(marker) rownames(classified_thresh_dat)[classified_thresh_dat[[marker]] == 1])
names(upset_dat) <- classification_markers

upset(fromList(upset_dat), number.angles=30, point.size=2, line.size=1, text.scale=1.2, nsets=length(upset_dat), 
      nintersects=30, order.by="freq", main.bar.color="#004586", sets.bar.color="#018F99")
```

#### Cell Type Quantification Tables

```{r}
meta_df <- cell_meta %>% filter(is_tumor)

create_summary <- function(meta_df) {
  cell_counts <- data.frame(cell_type=c("Epithelial", "T-Helper", "Cytotoxic T", "Macrophage", "B Cell", "Unknown"))
  cell_counts$num_cells <- sapply(cell_counts$cell_type, function(x) sum(meta_df$cell_type == x))
  
  summary_df <- data.frame(t(cell_counts$num_cells))
  colnames(summary_df) <- cell_counts$cell_type
  summary_df <- summary_df %>% mutate(Sample=params$sample_name,
                                      `T Cell`=`Cytotoxic T` + `T-Helper`,
                                      `Total Immune`=sum(across(c("Macrophage", "T Cell", "B Cell"))),
                                      `Total Classified`=sum(across(c("Total Immune", "Epithelial"))),
                                      `Total Cells`=sum(across(c("Total Classified", "Unknown"))),
                                      `Perc Classified`=100*`Total Classified`/`Total Cells`,
                                      `Perc Immune`=100*`Total Immune`/`Total Classified`,
                                      `Perc Epithelial`=100*Epithelial/`Total Classified`,
                                      `Perc Macrophage`=100*Macrophage/`Total Immune`,
                                      `Perc T Cell`=100*`T Cell`/`Total Immune`,
                                      `Perc T-Helper`=100*`T-Helper`/`Total Immune`,
                                      `Perc Cytotoxic T`=100*`Cytotoxic T`/`Total Immune`,
                                      `Perc B Cell`=100*`B Cell`/`Total Immune`)
  
  # Reorder columns
  summary_df %>% select(Sample, `Total Cells`, `Total Classified`, `Perc Classified`, `Total Immune`, `Perc Immune`, 
                        Epithelial, `Perc Epithelial`, Macrophage, `Perc Macrophage`, `B Cell`, `Perc B Cell`, 
                        `T Cell`, `Perc T Cell`, `Cytotoxic T`, `Perc Cytotoxic T`, `T-Helper`, `Perc T-Helper`)
}

# Tumor area
tumor_summary <- create_summary(cell_meta %>% filter(is_tumor))
tumor_summary %>% select(!c(`T Cell`, `Perc T Cell`)) %>% 
  mutate(across(starts_with("Perc"), function(col) paste0(round(col, 2), "%"))) %>% 
  t() %>% kable(caption="Tumor Area: Cell Type")
tumor_subtype_summary <- cell_meta %>% filter(is_tumor) %>% group_by(cell_subtype) %>% summarize(num_cells=n_distinct(cell_id))
tumor_subtype_summary %>% kable(caption="Tumor Area: Cell Subtype", col.names=c())

# Cell counts for PD-L1 scores
tumor_summary$`Total Classified PD-L1` <- cell_meta %>% filter(is_tumor, cell_type != "Unknown", pdl1_pos) %>% nrow()
tumor_summary$`Total Epithelial PD-L1` <- cell_meta %>% filter(is_tumor, cell_type == "Epithelial", pdl1_pos) %>% nrow()

if ("pdl2_pos" %in% colnames(cell_meta)) {
  tumor_summary$`Total Classified PD-L2` <- cell_meta %>% filter(is_tumor, cell_type != "Unknown", pdl2_pos) %>% nrow()
  tumor_summary$`Total Epithelial PD-L2` <- cell_meta %>% filter(is_tumor, cell_type == "Epithelial", pdl2_pos) %>% nrow()
}

# Adjacent normal (if any)
normal_summary <- create_summary(cell_meta %>% filter(!is_tumor))
normal_summary %>% select(!c(`T Cell`, `Perc T Cell`)) %>% 
  mutate(across(starts_with("Perc"), function(col) paste0(round(col, 2), "%"))) %>% 
  t() %>% kable(caption="Normal Area: Cell Type")
normal_subtype_summary <- cell_meta %>% filter(!is_tumor) %>% group_by(cell_subtype) %>% summarize(num_cells=n_distinct(cell_id))
normal_subtype_summary %>% kable(caption="Normal Area: Cell Subtype", col.names=c())

# Tumor glands
gland_summary <- create_summary(cell_meta %>% filter(is_gland))
gland_summary %>% select(!c(`T Cell`, `Perc T Cell`)) %>% 
  mutate(across(starts_with("Perc"), function(col) paste0(round(col, 2), "%"))) %>% 
  t() %>% kable(caption="Tumor Glands: Cell Type")
```

```{r}
# Save cell quantification results for tumor area and glands
write.csv(tumor_summary, paste0(params$results_outdir, "/", params$sample_name, "_tumor_area_cell_type_counts.csv"), row.names=FALSE, quote=FALSE)
write.csv(gland_summary, paste0(params$results_outdir, "/", params$sample_name, "_tumor_gland_cell_type_counts.csv"), row.names=FALSE, quote=FALSE)

tumor_subtype_summary_row <- tumor_subtype_summary$num_cells %>% t() %>% as.data.frame()
colnames(tumor_subtype_summary_row) <- tumor_subtype_summary$cell_subtype
tumor_subtype_summary_row <- cbind(Sample=params$sample_name, tumor_subtype_summary_row)
write.csv(tumor_subtype_summary_row, paste0(params$results_outdir, "/", params$sample_name, "_tumor_area_cell_subtype_counts.csv"), 
          row.names=FALSE, quote=FALSE)
```

```{r}
# Save filtered data with cell annotations/coordinates for plots
cell_meta <- merge(cell_meta, df %>% select(`Object ID`, `Centroid X µm`, `Centroid Y µm`) %>% rename(cell_id=`Object ID`), by="cell_id")
write.csv(cell_meta, paste0(params$results_outdir, "/", params$sample_name, "_filtered_cells.csv"), row.names=FALSE)
```
