---
title: "Figure 2"
output: html_document
---

<style type="text/css">
.main-container {
max-width: 1600px;
margin-left: 20px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
# knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(readxl)
library(dplyr)
library(knitr)
library(ggplot2)

quant_dir <- "quantification/"
```

```{r}
type_palette <- c(`Adenoid Cystic Carcinoma - Head & Neck`="#35B7CE", 
                  `Adenoid Cystic Carcinoma - Breast`="#C6EDF2", 
                  `Adenoid Cystic Carcinoma - Lung`="#91CCC1",  
                  `Adenoid Cystic Carcinoma - Metastatic`= "#0888A3", 
                  `Squamous Cell Carcinoma - Head & Neck`="#414487",
                  `Basal-Like Carcinoma - Breast`="#932667")
```             

```{r}
area_df <- read.csv(paste0(quant_dir, "tumor_area_cell_type_counts.csv"), check.names=FALSE)
```

```{r}
gland_df <- read.csv(paste0(quant_dir, "tumor_gland_cell_type_counts.csv"), check.names=FALSE)
gland_df <- gland_df %>% mutate(perc_ratio_intratumoral_t=100*`T Cell`/Epithelial)
```

```{r}
subtype_df <- read.csv(paste0(quant_dir, "tumor_area_cell_subtype_counts.csv"), check.names=FALSE)
subtype_df$Sample <- subtype_df$new_samp
```

```{r}
# Count total T-Helper including CK+ (appear to be valid, just overlapping epithelial cells)
subtype_df$`Total T-Helper` <- rowSums(subtype_df %>% select(starts_with("T-Helper")))

# Count total T-Regs including CK+
subtype_df$`Total T-Helper: Regulatory` <- rowSums(subtype_df %>% select(starts_with("T-Helper: Regulatory")))

# Count total cytotoxic Ts including CK+
subtype_df$`Total Cytotoxic T` <- rowSums(subtype_df %>% select(starts_with("Cytotoxic T")))

# Count total proliferating cytotoxic Ts including CK+
subtype_df$`Total Cytotoxic T: Proliferating` <- rowSums(subtype_df %>% select(starts_with("Cytotoxic T: Proliferating")))

# Count total epithelial
subtype_df$`Total Epithelial` <- rowSums(subtype_df %>% select(starts_with("Epithelial")))

subtype_df <- subtype_df %>% mutate(
  perc_t_reg_tot=100*`Total T-Helper: Regulatory`/`Total Classified`, 
  perc_t_reg_immune=100*`Total T-Helper: Regulatory`/`Total Immune`, 
  perc_t_reg_t_helper=100*`Total T-Helper: Regulatory`/`Total T-Helper`,
  perc_pro_t_cyto_tot_t_cyto=100*`Total Cytotoxic T: Proliferating`/`Total Cytotoxic T`,
  perc_pro_epi_tot_epi=100*`Epithelial: Proliferating`/`Total Epithelial`)
```

### Percentage Total Immune Cells Out of Total Cells

#### Tumor Area

```{r, fig.width=14, fig.height=6}
dat <- area_df
dat$perc_immune_label <- paste0(round(dat$`Perc Immune`, 1), "%")
dat$tot_immune_label <- dat$`Total Immune`

ggplot(dat, aes(x=Sample, y=`Perc Immune`, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_immune_label, x=Sample, y=`Perc Immune`+0.8), size=4, color="gray20", fontface="bold") +
  geom_text(aes(label=tot_immune_label), position=ggpp::position_stacknudge(vjust=0.5, y=0), size=4) +
  scale_y_continuous(expand=c(0, 1)) +
  # guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold"),
        legend.title=element_text(size=16, face="bold")) +
  labs(y=" \nImmune Cells/Total Cells [%]", x="")
```

#### Tumor Glands

```{r, fig.width=14, fig.height=6}
dat <- gland_df
dat$perc_immune_label <- paste0(round(dat$`Perc Immune`, 1), "%")
dat$tot_immune_label <- dat$`Total Immune`

ggplot(dat, aes(x=Sample, y=`Perc Immune`, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_immune_label, x=Sample, y=`Perc Immune`+0.6), size=4, color="gray20", fontface="bold") +
  geom_text(aes(label=tot_immune_label), position=ggpp::position_stacknudge(vjust=0.5, y=0), size=4) +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y="Immune Cells/Total Cells\nwithin Malignant Epithelium [%]", x="")
```

### Percentage Macrophage/T-Helper/Cytotoxic T/B Cells Out of Total Immune Cells

```{r, fig.width=10, fig.height=9}
dat <- area_df %>% 
  tidyr::pivot_longer(c(`Perc B Cell`, `Perc T-Helper`, `Perc Cytotoxic T`, `Perc Macrophage`), 
                      names_to="Cell Type", values_to="Percent of Total Immune Cells") %>% 
  mutate(`Cell Type`=case_when(`Cell Type` == "Perc Macrophage" ~ "Macrophage",
                               `Cell Type` == "Perc T-Helper" ~ "Helper T Cell",
                               `Cell Type` == "Perc Cytotoxic T" ~ "Cytotoxic T Cell",
                               `Cell Type` == "Perc B Cell" ~ "B Cell"))

ggplot(dat %>% mutate(Sample=factor(Sample, levels=rev(levels(Sample)))), 
       aes(x=`Percent of Total Immune Cells`, y=Sample, fill=`Cell Type`)) + 
  geom_col(alpha=0.75) +
  scale_fill_manual(values=c(`Helper T Cell`="#FE9F6D", `Cytotoxic T Cell`="#DD513A", 
                             `Macrophage`="#4BAD9F", `B Cell`="#2F6B8E"),
                    breaks=c("Macrophage", "Helper T Cell", "Cytotoxic T Cell", "B Cell")) +
  theme_minimal() +
  theme(text=element_text(size=18, color="black", face="bold"),
        axis.text=element_text(size=14), 
        axis.title.x=element_text(margin=unit(c(4, 0, 0, 0), "mm"), size=14), 
        legend.text=element_text(size=12, face="plain"),
        legend.title=element_text(size=14),
        legend.direction="horizontal", legend.position="top",
        legend.margin=margin(0, 0, -3, 0)) +
  scale_x_continuous(expand=c(0, 1.2), n.breaks=20) +
  labs(x="% of Total Immune Cells", y="")
```

### Ratio Intratumoral T-Cells To Cancer Cells

```{r, fig.width=14, fig.height=6}
dat <- gland_df
dat$perc_ratio_t_label <- paste0(round(dat$perc_ratio_intratumoral_t, 2), "%")

ggplot(dat, aes(x=Sample, y=perc_ratio_intratumoral_t, fill=`Cancer Type`)) + 
  geom_col(alpha=0.75, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_ratio_t_label, x=Sample, y=perc_ratio_intratumoral_t+0.3), size=3.8, color="gray20", fontface="bold") +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y="T-Cells/Total Cells\nwithin Malignant Epithelium [%]", x="")
```

### PDL1 Score

```{r}
pdl1_df <- read.csv(paste0(quant_dir, "tumor_area_PDL1.csv"))
```

```{r, fig.width=14, fig.height=6}
dat <- pdl1_df
dat <- dat %>% tidyr::pivot_longer(c(CPS, TPS), names_to="Score Type", values_to="PDL1 Score")
dat$CPS_star <- ifelse(dat$`Score Type` == "CPS" & dat$`PDL1 Score` > 1, "*", "")
dat$TPS_star <- ifelse(dat$`Score Type` == "TPS" & dat$`PDL1 Score` > 1, "*", "")
dat$label <- ifelse(dat$`PDL1 Score` > 1, paste("*", round(dat$`PDL1 Score`, 2)), "")

ggplot(dat, aes(x=Sample, y=`PDL1 Score`, fill=`Score Type`)) + 
  geom_col(position=position_dodge2(), width=0.7, alpha=0.75) +
  scale_fill_manual(values=c(CPS="#9AD8DE", TPS="#388DBA")) +
  geom_hline(yintercept=1, linetype="dashed", linewidth=1, color="#94221F") +
  geom_text(aes(label=CPS_star, x=as.integer(Sample) - 0.17, y=`PDL1 Score`+1), size=6, color="black", fontface="bold") +
  geom_text(aes(label=TPS_star, x=as.integer(Sample) + 0.17, y=`PDL1 Score`+1), size=6, color="black", fontface="bold") +
  theme_minimal() +
  guides(fill="none") +
  theme(text=element_text(size=16, color="black", face="bold"), axis.text=element_text(size=18), 
        axis.text.x=element_text(angle=60, hjust=1),
        axis.title.y=element_text(size=18, margin=unit(c(0, 4, 0, 0), "mm"))) +
  labs(y="PDL1 Score", x="")
```

___
# Supplemental Figures

### Percent Ratio Proliferating Cytotoxic T / Total Cytotoxic T

```{r, fig.width=16, fig.height=6}
dat <- subtype_df
dat$perc_label <- paste0(round(dat$perc_pro_t_cyto_tot_t_cyto, 1), "%")

ggplot(dat, aes(x=Sample, y=perc_pro_t_cyto_tot_t_cyto, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_label, x=Sample, y=perc_pro_t_cyto_tot_t_cyto+1.7), size=4, color="black", fontface="bold") +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y="Proliferating Cytotoxic T Cells /\nTotal Cytotoxic T Cells [%]", x="")
```

### Percent Ratio T-Regs / Total Cells

```{r, fig.width=16, fig.height=6}
dat$perc_label <- paste0(round(dat$perc_t_reg_tot, 1), "%")

ggplot(dat, aes(x=Sample, y=perc_t_reg_tot, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_label, x=Sample, y=perc_t_reg_tot+0.13), size=4, color="black", fontface="bold") +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y=" \nRegulatory T Cells / Total Cells [%]", x="")
```

### Percent Ratio T-Regs / Total Immune Cells

```{r, fig.width=16, fig.height=6}
dat$perc_label <- paste0(round(dat$perc_t_reg_immune, 1), "%")

ggplot(dat, aes(x=Sample, y=perc_t_reg_immune, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_label, x=Sample, y=perc_t_reg_immune+0.6), size=4, color="black", fontface="bold") +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y=" \nRegulatory T Cells / Total Immune Cells [%]", x="")
```

### Percent Ratio T-Regs / Total T-Helper

```{r, fig.width=16, fig.height=6}
dat$perc_label <- paste0(round(dat$perc_t_reg_t_helper, 1), "%")

ggplot(dat, aes(x=Sample, y=perc_t_reg_t_helper, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_label, x=Sample, y=perc_t_reg_t_helper+1.5), size=4, color="black", fontface="bold") +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y="Regulatory T-Helper Cells /\nTotal T-Helper Cells [%]", x="")
```

### Percent Proliferating Epithelial Cells / Total Epithelial Cells

```{r, fig.width=16, fig.height=6}
dat$perc_label <- paste0(round(dat$perc_pro_epi_tot_epi, 1), "%")

ggplot(dat, aes(x=Sample, y=perc_pro_epi_tot_epi, fill=`Cancer Type`)) + 
  geom_col(alpha=0.7, width=0.95) +
  scale_fill_manual(values=type_palette) +
  geom_text(aes(label=perc_label, x=Sample, y=perc_pro_epi_tot_epi+2), size=4, color="black", fontface="bold") +
  guides(fill="none") +
  theme_minimal() +
  theme(text=element_text(size=16, color="black"),
        axis.text=element_text(size=18, face="bold"),
        axis.text.x=element_text(angle=60, hjust=1, face="bold"),
        axis.title.y=element_text(size=16, margin=unit(c(0, 4, 0, 0), "mm"), face="bold")) +
  labs(y="Proliferating Epithelial Cells /\nTotal Epithelial Cells [%]", x="")
```

### PDL2 Score

```{r}
pdl2_df <- read.csv(paste0(quant_dir, "tumor_area_PDL2.csv"))

# PD-L2 data only available for some samples
pdl2_df <- pdl2_df[!is.na(pdl2_df$CPS),]

pdl2_df$Sample <- factor(pdl2_df$Sample, level=levels(pdl2_df$Sample)[levels(pdl2_df$Sample) %in% pdl2_df$Sample])
```

```{r, fig.width=16, fig.height=6}
dat <- pdl2_df
dat <- dat %>% tidyr::pivot_longer(c(CPS, TPS), names_to="Score Type", values_to="PDL2 Score")
dat$CPS_star <- ifelse(dat$`Score Type` == "CPS" & dat$`PDL2 Score` > 1, "*", "")
dat$TPS_star <- ifelse(dat$`Score Type` == "TPS" & dat$`PDL2 Score` > 1, "*", "")
dat$label <- ifelse(dat$`PDL2 Score` > 1, paste("*", round(dat$`PDL2 Score`, 2)), "")

ggplot(dat, aes(x=Sample, y=`PDL2 Score`, fill=`Score Type`)) + 
  geom_col(position=position_dodge2(), width=0.7, alpha=0.75) +
  scale_fill_manual(values=c(CPS="#9AD8DE", TPS="#388DBA")) +
  geom_hline(yintercept=1, linetype="dashed", linewidth=1, color="#94221F") +
  geom_text(aes(label=CPS_star, x=as.integer(Sample) - 0.17, y=`PDL2 Score`+1), size=6, color="black", fontface="bold") +
  geom_text(aes(label=TPS_star, x=as.integer(Sample) + 0.17, y=`PDL2 Score`+1), size=6, color="black", fontface="bold") +
  theme_minimal() +
  scale_y_continuous(limits=c(0, 100)) +
  guides(fill="none") +
  theme(text=element_text(size=16, color="black", face="bold"), axis.text=element_text(size=18), 
        axis.text.x=element_text(angle=60, hjust=1),
        axis.title.y=element_text(size=18, margin=unit(c(0, 4, 0, 0), "mm"))) +
  labs(y="PDL2 Score", x="")
```
